A Memory-Bounded, Deterministic and Terminating Semantics for the Synchronous Programming Language~\CEU

### 2018-04-15 01-intro.tex

\begin{description}
\item[Memory Boundedness:]
At compile time, we can ensure that the program fits in the device's restricted
memory and that it will not grow unexpectedly during runtime.
\item[Deterministic Execution:]
We can simulate an entire program execution providing an input history with the
guarantee that it will always have the same behavior.
This can be done in negligible time in a controlled development environment
before deploying the application in the actual device (e.g., by multiple
developers in standard desktops).
\item[Terminating Reactions:]
Real-time applications must guarantee responses within specified deadlines.
A terminating semantics enforces upper bounds for all reactions and guarantees
that programs always progress with the time.
%In addition, it also ensures that programs always reach an idle state amenable
%to entering in standby mode, which can improve the autonomy of battery-powered
%devices.
\end{description}

>>>>

\begin{description}
\item[Memory Boundedness:]
At compile time, we can ensure that the program fits in the device's restricted
memory and that it will not grow unexpectedly during runtime.
\item[Deterministic Execution:]
We can simulate an entire program execution providing an input history with the
guarantee that it will always have the same behavior.
This can be done in negligible time in a controlled development environment
before deploying the application to the actual device (e.g., by multiple
developers in standard desktops).
\item[Terminating Reactions:]
Real-time applications must guarantee responses within specified deadlines.
A terminating semantics enforces upper bounds for all reactions and guarantees
that programs always progress with the time.
\end{description}

### 2018-04-15 02-ceu.tex

\begin{lstlisting}[
  numbers=left,
  basicstyle=\ttfamily\footnotesize,
  float=h,
  caption={A program in \CEU that toggles a LED on every radio receive,
           terminating on a button press always with the LED off.},
  label={lst.ceu},
]
input void BUTTON;
input void RADIO_RECV;
par/or do
    await BUTTON;
with
    finalize with
        _led(0);
    end
    loop do
        _led(1);
        await RADIO_RECV;
        _led(0);
        await RADIO_RECV;
    end
end
\end{lstlisting}

In \CEU, any identifier prefixed with an underscore (e.g., \code{\_led}) is
passed unchanged to the underlying~C compiler.
%
Therefore, access to~C is straightforward and syntactically traceable.
%
To ensure that programs operate under the synchronous hypothesis, the compiler
environment should only provide access to~C operations that can be assumed to
be instantaneous, such as non-blocking~I/O and simple data structure accessors.%
\footnote {
The actual implementation of Céu supports a command-line option that accepts a
white list of library calls.
If a program tries to use a function not in this list, the compiler raises an
error.
}

@@

\begingroup
\begin{figure}[h]
\centering
\begin{tikzpicture}[x=.9em,y=.9em,
  font=\footnotesize\sffamily,>=stealth,
  time/.style={font=\scriptsize\sffamily},
  cross/.style={inner sep=0pt,outer sep=0pt},
  evt/.style={fill=white,inner sep=1.85pt,outer sep=0pt},
  rng/.style={fill=white,inner sep=1pt,outer sep=0pt,
    font=\ttfamily\scriptsize}]
  \draw[->]
  (0,0)
    node[align=right,anchor=east,xshift=-.5em]
        {Continuous\\Timeline\\(Real World)}
    node[anchor=north west,xshift=-.5em,time]
        {0ms000us}
    node[anchor=north west,xshift=8.75em,time]
        {17ms521us}
  -- (18.5,0)
    node[anchor=north east,xshift=.1em,time]
        {30ms000us};
  \node[cross] at (0,0){x};
  %%
  \coordinate(Ext) at ($(0,0)!.5!(0,-5)$);
  \coordinate(A) at (Ext-|4,0);
  \coordinate(B) at (Ext-|5.5,0);
  \coordinate(C) at (Ext-|9,0);
  \coordinate(D) at (Ext-|15,0);
  %%
  \draw[semithick,dotted] (0,1-|A) -- (0,-6-|A);
  \draw[semithick,dotted] (0,1-|B) -- (0,-6-|B);
  \draw[semithick,dotted] (0,1-|C) -- (0,-6-|C);
  \draw[semithick,dotted] (0,1-|D) -- (0,-5-|D);
  \node[evt](Aevt) at (A) {\textbf{A}};
  \node[evt](Bevt) at (B) {\textbf{B}};
  \node[evt](Cevt) at (C) {\textbf{C}};
  \node[evt](Devt) at (D) {\textbf{D}};
  \node[anchor=east,xshift=-.5em] at (A){\emph{External Input}};
  %%
  \draw[-]
  (0,-5)
    node[align=right,anchor=east,xshift=-.5em]
        {Discrete\\Timeline\\(\CEU)}
  -- (18.5,-5);
  \node[cross](P0) at (0,-5){x};
  \node[cross](P1) at (4.75,-5){x};
  \node[cross](P2) at (10,-5){x};
  \node[cross](P3) at (12.5,-5){x};
  \node[cross](P4) at (14.5,-5){x};
  \node[cross](P5) at (15.5,-5){x};
  \node[cross](P6) at (18.5,-5){x};
  \node[rng] at ($(P0)!.5!(P1)$){boot-0};
  \node[rng] at ($(P1)!.5!(P2)$){A-1};
  \node[rng] at ($(P2)!.5!(P3)$){B-2};
  \node[rng] at ($(P3)!.5!(P4)$){C-3};
  \node[rng] at ($(P5)!.5!(P6)$){D-4};
  \draw[decorate,decoration={brace,raise=-1pt,mirror,amplitude=2pt}]
  ($(P4)+(0,-.5)$) -- node[below,time] {Idle period} ($(P5)+(0,-.5)$);
  %%
  \begin{scope}[shorten >=1.5pt]
  \draw[->] (Aevt) -- (P1);
  \draw[->] (Bevt) -- (P2);
  \draw[->] (Cevt) -- (P3);
  \draw[->] (Devt) -- (P5);
  \end{scope}
\end{tikzpicture}
%\includegraphics[width=\columnwidth]{tick_min}
\caption{The discrete notion of time in \CEU.}
\label{fig.ticks}
\end{figure}
\endgroup

@@

In example \ax, the two assignments to \code{x} can only execute in
reactions to different events \code{A} and \code{B} (ln.~5, 8), which cannot
occur simultaneously by definition.
Hence, for the sequence of events \code{A->B}, \code{x} becomes \code{4}
(\code{(1+1)*2}), while for \code{B->A}, \code{x} becomes \code{3}
(\code{(1*2)+1}).

In example \bx, the two assignments to \code{y} are simultaneous because
they execute in reaction to the same event \code{A}.
Since \CEU employs lexical order for intra-reaction statements, the execution
is still deterministic, and \code{y} always becomes \code{4} (\code{(1+1)*2}).
%
However, note that an apparently innocuous change in the order of trails
modifies the behavior of the program.
%
To mitigate this threat, \CEU performs concurrency checks at compile time to
detect conflicting accesses to shared variables:
if a variable is written in a trail segment, then a concurrent trail segment
cannot access that variable~\cite{ceu.sensys13}.
%
Nonetheless, the static checks are optional and are not a prerequisite for the
deterministic semantics of the language.

@@

\begin{figure}[h]
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}[numbers=right,basicstyle=\ttfamily\footnotesize]
par/or do
   var _msg_t msg;
   <...> // prepare msg
   finalize
      _send(&msg);
   with
      _cancel(&msg);
   end
   await SEND_ACK;
with
   <...>
end
//
\end{lstlisting}
\small{\ax\!Local resource finalization\rlap{.}}
\end{minipage}%
%
\begin{minipage}[t]{0.53\linewidth}
\begin{lstlisting}[xleftmargin=2em,basicstyle=\ttfamily\footnotesize]
par/or do
   var _FILE* f;
   finalize
      f = _fopen(...);
   with
      _fclose(f);
   end
   _fwrite(..., f);
   await A;
   _fwrite(..., f);
with
   <...>
end
\end{lstlisting}
\small{\quad\bx\!External resource finalization\rlap{.}}
\end{minipage}%
%\rule{8.4cm}{0.37pt}
\caption{%
\CEU enforces the use of finalization to prevent dangling pointers and memory leaks.}
\label{lst.fin.ceu}
\end{figure}

The \code{par/or} of \CEU is an orthogonal abortion mechanism because the two
sides in the composition need not be tweaked with synchronization primitives nor
state variables to affect each other.
%
In addition, abortion is \emph{immediate} in the sense that it executes
atomically in the current micro reaction.
%
Immediate orthogonal abortion is a distinctive feature of synchronous languages
and cannot be expressed effectively in traditional (asynchronous)
multi-threaded languages~\cite{esterel.preemption,sync_async.threadsstop}.

@@

\CEU also enforces the use of \code{finalize} for system calls that deal with
pointers representing resources, as illustrated in the two examples of
Figure~\ref{lst.fin.ceu}.
%
%\begin{itemize}
 If \CEU \textbf{passes} a pointer to a system call (ln.~\ax:5), the
          pointer represents a \textbf{local} resource (ln.~\ax:2) that
          requires finalization (ln.~\ax:7).
 If \CEU \textbf{receives} a pointer from a system call return
          (ln.~ \bx:4), the pointer represents an \textbf{external} resource
          (ln.~\bx:2) that requires finalization (ln.~\bx:6).
%\end{itemize}

\CEU tracks the interaction of system calls with pointers and requires
finalization clauses to accompany them.
%
In Figure~\ref{lst.fin.ceu}.a, the local variable \code{msg}
(ln.~2) is an internal resource passed as a pointer to \code{\_send} (ln.~5),
which is an asynchronous call that transmits the buffer in the background.
If the block aborts (ln.~11) before receiving an acknowledge from the
environment (ln.~9), the local \code{msg} goes out of scope and the external
transmission now holds a \emph{dangling pointer}.
The finalization ensures that the transmission also aborts (ln.~7).
%
In Figure~\ref{lst.fin.ceu}.b, the call to \code{\_fopen} (ln.
4) returns an external file resource as a pointer.
If the block aborts (ln.~12) during the \code{await A} (ln.~9), the file
remains open as a \emph{memory leak}.
The finalization ensures that the file closes properly (ln.~6).
%
In both cases, the code does not compile without the \code{finalize}
construct.%
\footnote{%
The compiler only forces the programmer to write the finalization clause, but
cannot check if it actually handles the resource properly.
}

>>>>

\begin{lstlisting}[
  numbers=left,
  basicstyle=\ttfamily\footnotesize,
  float=h,
  caption={A \CEU program that toggles a LED on radio receive,
           terminating on a button press always with the LED~off.},
  label={lst.ceu},
]
input void BUTTON;
input void RADIO_RECV;
par/or do
    await BUTTON;
with
    finalize with
        _led(0);
    end
    loop do
        _led(1);
        await RADIO_RECV;
        _led(0);
        await RADIO_RECV;
    end
end
\end{lstlisting}

In \CEU, any identifier prefixed with an underscore (e.g., \code{\_led}) is
passed unchanged to the underlying~C compiler.
%
Therefore, access to~C is straightforward and syntactically traceable.
%
To ensure that programs operate under the synchronous hypothesis, the compiler
environment should only provide access to~C operations that can be assumed to
be instantaneous, such as non-blocking~I/O and simple data structure accessors.%
\footnote {
The actual implementation of Céu supports a command-line option that accepts a
white list of library calls.
If a program tries to use a function not in the list, the compiler raises an
error.
}

@@

\begingroup
\begin{figure}[t]
\centering
\begin{tikzpicture}[x=.9em,y=.9em,
  font=\footnotesize\sffamily,>=stealth,
  time/.style={font=\scriptsize\sffamily},
  cross/.style={inner sep=0pt,outer sep=0pt},
  evt/.style={fill=white,inner sep=1.85pt,outer sep=0pt},
  rng/.style={fill=white,inner sep=1pt,outer sep=0pt,
    font=\ttfamily\scriptsize}]
  \draw[->]
  (0,0)
    node[align=right,anchor=east,xshift=-.5em]
        {Continuous\\Timeline\\(Real World)}
    node[anchor=north west,xshift=-.5em,time]
        {0ms000us}
    node[anchor=north west,xshift=8.75em,time]
        {17ms521us}
  -- (18.5,0)
    node[anchor=north east,xshift=.1em,time]
        {30ms000us};
  \node[cross] at (0,0){x};
  %%
  \coordinate(Ext) at ($(0,0)!.5!(0,-5)$);
  \coordinate(A) at (Ext-|4,0);
  \coordinate(B) at (Ext-|5.5,0);
  \coordinate(C) at (Ext-|9,0);
  \coordinate(D) at (Ext-|15,0);
  %%
  \draw[semithick,dotted] (0,1-|A) -- (0,-6-|A);
  \draw[semithick,dotted] (0,1-|B) -- (0,-6-|B);
  \draw[semithick,dotted] (0,1-|C) -- (0,-6-|C);
  \draw[semithick,dotted] (0,1-|D) -- (0,-5-|D);
  \node[evt](Aevt) at (A) {\textbf{A}};
  \node[evt](Bevt) at (B) {\textbf{B}};
  \node[evt](Cevt) at (C) {\textbf{C}};
  \node[evt](Devt) at (D) {\textbf{D}};
  \node[anchor=east,xshift=-.5em] at (A){\emph{External Input}};
  %%
  \draw[-]
  (0,-5)
    node[align=right,anchor=east,xshift=-.5em]
        {Discrete\\Timeline\\(\CEU)}
  -- (18.5,-5);
  \node[cross](P0) at (0,-5){x};
  \node[cross](P1) at (4.75,-5){x};
  \node[cross](P2) at (10,-5){x};
  \node[cross](P3) at (12.5,-5){x};
  \node[cross](P4) at (14.5,-5){x};
  \node[cross](P5) at (15.5,-5){x};
  \node[cross](P6) at (18.5,-5){x};
  \node[rng] at ($(P0)!.5!(P1)$){boot-0};
  \node[rng] at ($(P1)!.5!(P2)$){A-1};
  \node[rng] at ($(P2)!.5!(P3)$){B-2};
  \node[rng] at ($(P3)!.5!(P4)$){C-3};
  \node[rng] at ($(P5)!.5!(P6)$){D-4};
  \draw[decorate,decoration={brace,raise=-1pt,mirror,amplitude=2pt}]
  ($(P4)+(0,-.5)$) -- node[below,time] {Idle period} ($(P5)+(0,-.5)$);
  %%
  \begin{scope}[shorten >=1.5pt]
  \draw[->] (Aevt) -- (P1);
  \draw[->] (Bevt) -- (P2);
  \draw[->] (Cevt) -- (P3);
  \draw[->] (Devt) -- (P5);
  \end{scope}
\end{tikzpicture}
%\includegraphics[width=\columnwidth]{tick_min}
\caption{The discrete notion of time in \CEU.}
\label{fig.ticks}
\end{figure}
\endgroup

@@

In \ax, the two assignments to \code{x} can only execute in
reactions to different events \code{A} and \code{B} (ln.~5, 8), which cannot
occur simultaneously by definition.
Hence, for the sequence of events \code{A->B}, \code{x} becomes \code{4}
(\code{(1+1)*2}), while for \code{B->A}, \code{x} becomes \code{3}
(\code{(1*2)+1}).

In \bx, the two assignments to \code{y} are simultaneous because
they execute in reaction to the same event \code{A}.
Since \CEU employs lexical order for intra-reaction statements, the execution
is still deterministic, and \code{y} always becomes \code{4} (\code{(1+1)*2}).
%
However, note that an apparently innocuous change in the order of trails
modifies the behavior of the program.
%
To mitigate this threat, \CEU performs concurrency checks at compile time to
detect conflicting accesses to shared variables:
if a variable is written in a trail segment, then a concurrent trail segment
cannot access that variable~\cite{ceu.sensys13}.
%
Nonetheless, the static checks are optional and are not a prerequisite for the
deterministic semantics of the language.

@@

The \code{par/or} of \CEU is an orthogonal abortion mechanism because the two
sides in the composition need not be tweaked with synchronization primitives nor
state variables to affect each other.
%
In addition, abortion is \emph{immediate} in the sense that it executes
atomically inside the current micro reaction.
%
Immediate orthogonal abortion is a distinctive feature of synchronous languages
and cannot be expressed effectively in traditional (asynchronous)
multi-threaded languages~\cite{esterel.preemption,sync_async.threadsstop}.

@@

\CEU also enforces, at compile time, the use of \code{finalize} for system
calls that deal with pointers representing resources, as illustrated in the two
examples of Figure~\ref{lst.fin.ceu}.
%
%\begin{itemize}
 If \CEU \textbf{passes} a pointer to a system call (ln.~\ax:5), the
          pointer represents a \textbf{local} resource (ln.~\ax:2) that
          requires finalization (ln.~\ax:7).
 If \CEU \textbf{receives} a pointer from a system call return
          (ln.~ \bx:4), the pointer represents an \textbf{external} resource
          (ln.~\bx:2) that requires finalization (ln.~\bx:6).
%\end{itemize}

\CEU tracks the interaction of system calls with pointers and requires
finalization clauses to accompany them.
%
In Figure~\ref{lst.fin.ceu}.a, the local variable \code{msg}
(ln.~2) is an internal resource passed as a pointer to \code{\_send} (ln.~5),
which is an asynchronous call that transmits the buffer in the background.
If the block aborts (ln.~11) before receiving an acknowledge from the
environment (ln.~9), the local \code{msg} goes out of scope and the external
transmission now holds a \emph{dangling pointer}.
The finalization ensures that the transmission also aborts (ln.~7).
%
In Figure~\ref{lst.fin.ceu}.b, the call to \code{\_fopen} (ln.
4) returns an external file resource as a pointer.
If the block aborts (ln.~12) during the \code{await A} (ln.~9), the file
remains open as a \emph{memory leak}.
The finalization ensures that the file closes properly (ln.~6).
%
In both cases, the code does not compile without the \code{finalize}
construct.%
\footnote{%
The compiler only forces the programmer to write the finalization clause, but
cannot check if it actually handles the resource properly.}

\begin{figure}[t]
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}[numbers=right,basicstyle=\ttfamily\footnotesize]
par/or do
   var _msg_t msg;
   <...> // prepare msg
   finalize
      _send(&msg);
   with
      _cancel(&msg);
   end
   await SEND_ACK;
with
   <...>
end
//
\end{lstlisting}
\small{\ax\!Local resource finalization\rlap{.}}
\end{minipage}%
%
\begin{minipage}[t]{0.53\linewidth}
\begin{lstlisting}[xleftmargin=2em,basicstyle=\ttfamily\footnotesize]
par/or do
   var _FILE* f;
   finalize
      f = _fopen(...);
   with
      _fclose(f);
   end
   _fwrite(..., f);
   await A;
   _fwrite(..., f);
with
   <...>
end
\end{lstlisting}
\small{\quad\bx\!External resource finalization\rlap{.}}
\end{minipage}%
%\rule{8.4cm}{0.37pt}
\caption{%
\CEU enforces the use of finalization to prevent dangling pointers and memory leaks.}
\label{lst.fin.ceu}
\end{figure}

### 2018-04-15 03-semantics.tex

The core of our semantics describes how a program reacts to a single external
input event, i.e., starting from an input event, how the program behaves and
becomes idle again to proceed to a subsequent reaction.
%
We use a set of small-step operational rules~\cite{Plotkin-G-D-1981} which
ensure that at most one transition is possible at any time, resulting in
deterministic reactions.
%
The transition rules map a triple with a program~$p$, a stack level~$n$, and an
emitted event~$e$ to a modified triple as follows:
\[
  \<p,n,e>\trans\<p',n',e'>\,,
\]
where~$p,p'\in\P$ are abstract-language programs, $n,n'\in\N$ are nonnegative
integers representing the current stack level, and~$e,e'\in\E\cup\{\nil\}$ are
the events emitted before and after the transition (both possibly being the
empty event~$\nil$).
%$\E$ is a set of primitive events.

>>>>

The core of our semantics describes how a program reacts to a single external
input event, i.e., starting from an input event, how the program behaves and
becomes idle again to proceed to a subsequent reaction.
%
We use a set of small-step operational rules~\cite{Plotkin-G-D-1981} which
ensure that at most one transition is possible at any time, resulting in
deterministic reactions.
%
The transition rules map a triple with a program~$p$, a stack level~$n$, and an
emitted event~$e$ to a modified triple as follows:
\[
  \<p,n,e>\trans\<p',n',e'>\,,
\]
where~$p,p'\in\P$ are abstract-language programs, $n,n'\in\N$ are non-negative
integers representing the current stack level, and~$e,e'\in\E\cup\{\nil\}$ are
the events emitted before and after the transition (both possibly being the
empty event~$\nil$).
%$\E$ is a set of primitive events.
